Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-05 paper.tex
Fixed TODOs and code listings

\documentclass{juliacon}
\setcounter{page}{1}

@@

DONE:
JuliaHealth is an open-source collective dedicated to bringing the Julia programming language to various avenues of health research domains such as observational health, medical imaging, standards interoperability, etc.
It leverages the strengths of the Julia language to introduce new approaches to these spaces, augment and improve existing workflows, and provide a central community structure for health researchers both part of and entering into the Julia ecosystem.
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

@@

TODO: Cite FunOHDSI
TODO: Cite OHDSICohortExpressions
TODO: Cite OHDSI Symposium poster on JuliaHealth
TODO: Register OHDSI Symposium poster to Zenodo
TODO: Cite JuliaCon 2023 presentation
TODO: Register JuliaCon 2023 presentation to Zenodo

@@

DONE:

@@

TODO: Add small example showing iterative process 
TODO: Reference capr somehow

@@

TODO: Add database connector example
TODO: Reference database connector example
TODO: Mention package extensions utility

@@

TODO: Add tutorial to separate repository
TODO: Create tutorial about Eunomia
TODO: Reference Eunomia

@@

\begin{verbatim}
  import HealthSampleData: Eunomia
  import OMOPCDMCohortCreator as OCC

  eunomia = Eunomia()
  conn = DB(eunomia)

@@

  occ.GenerateTables(conn)
\end{verbatim}

TODO: Replace with DBConnector syntax

@@

\begin{verbatim}
  using JSON3 
  using OHDSICohortExpressions

@@

  model = Model(cdm_version=v"5.3.1", cdm_schema="main",
                       vocabulary_schema="main", results_schema="main",
                       target_schema="main", target_table="cohort");

  sql = translate(cohort_expression, dialect=:sqlite, model=model, cohort_definition_id=1);

  [DBI.execute(conn, sub_query) for sub_query in split(sql, ";")[1:end-1]]
\end{verbatim}

@@

TODO: Clean up the code a little nicer
TODO: Make sure the JSON3 reading of a cohort works properly

@@

\begin{verbatim}
  patients = DBI.execute(conn, "SELECT subject_id AS person_id FROM cohort;") |> DF.DataFrame
\end{verbatim}

TODO: Replace this line from Jay's PR for getting cohorts

\begin{verbatim}
  patients_race = occ.GetPatientRace(patients.person_id, conn)
  patients_gender = occ.GetPatientGender(patients.person_id, conn)

  patients_age_group = occ.GetPatientAgeGroup(
    patients.person_id, 
    conn; 
    age_groupings = [[age, age + 4] for age in 0:5:119]
  )
\end{verbatim}

@@

\begin{verbatim}
  import DataFrames as DF

  patients_characterized = DF.outerjoin(patients_race,
                                              patients_gender,
                                              patients_age_group;
                                              on = :person_id, 
                                              matchmissing = :equal)
  patients_characterized = patients_characterized[:, DF.Not(:person_id)]
  patient_groups = DF.groupby(patients_characterized, [:race_concept_id, :gender_concept_id, :age_group])
  patient_groups = DF.combine(patient_groups, DF.nrow => :count)
\end{verbatim}

@@

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort Count & Total Count & Prev. (\%) \\ \hline
        Black or African American & Female & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        White & Male & 70 - 74 & 68 & 96 & 70.83 \\ \hline
        White & Male & 45 - 49 & 47 & 74 & 63.51 \\ \hline
        White & Female & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        Black or African American & Female & 55 - 59 & 7 & 19 & 36.84 \\ \hline
        White & Male & 75 - 79 & 45 & 56 & 80.35 \\ \hline
        White & Female & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        White & Female & 45 - 49 & 41 & 92 & 44.56 \\ \hline
        Asian & Female & 80 - 84 & 4 & 5 & 80.0  \hline
        Asian & Male & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        Black or African American & Female & 45 - 49 & 10 & 23 & 43.47 \\ \hline
        No matching concept & Female & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        Black or African American & Male & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
\end{table}

@@

TODO: Check that I equated currying correctly here.

\begin{verbatim}
import Base:
  Fix2

CGetPatientRace = Base.Fix2(occ.GetPatientRace, conn)
CGetPatientGender = Base.Fix2(occ.GetPatientGender, conn)
CGetPatientAgeGroup = Base.Fix2(occ.GetPatientAgeGroup, conn)
\end{verbatim}

@@

\begin{verbatim}
import Chain:
  @chain

CofactorPatients(patients) = @chain patients begin
    CGetPatientRace_fixed
    CGetPatientGender_fixed
    CGetPatientAgeGroup_fixed
end
\end{verbatim}

\begin{verbatim}
CharacterizePatients(patients) = @chain patients begin
    _[:, DF.Not(:person_id)]
    DF.groupby(_, names(_))
    DF.combine(_, DF.nrow => :count)
    occ.ExecuteAudit
end
\end{verbatim}

@@

\begin{verbatim}
RunStudy(patients) = @chain patients begin
    CofactorPatients
    CharacterizePatients
end
\end{verbatim}

@@

\begin{verbatim}
import Distributed:
  @distributed

@@

vcat(result...) |> audit_patient_groups
\end{verbatim}

@@

\begin{verbatim}
  GetPatientAgeGroup([1]; age_groupings = [[0, 40], [41, 80]])
\end{verbatim}

\begin{verbatim}
  SELECT
    "PERSON_2"."person_id",
    (CASE 
      WHEN ("PERSON_2"."age" < 41) 
        THEN '0 - 40' 
      WHEN ("PERSON_2"."age" < 81) 
        THEN '41 - 80' END) AS "age_group"
  FROM (
    SELECT
      "PERSON_1"."person_id",
      (2023 - "PERSON_1"."year_of_birth") AS "age"
    FROM "PERSON" AS "PERSON_1"
    WHERE ("PERSON_1"."person_id" IN (1))
  ) AS "PERSON_2"
\end{verbatim}

>>>>

\documentclass{juliacon}

\usepackage{listings}

\setcounter{page}{1}

@@

% DONE:
JuliaHealth is an open-source collective dedicated to bringing the Julia programming language to various avenues of health research domains such as observational health, medical imaging, standards interoperability, etc.
It leverages the strengths of the Julia language to introduce new approaches to these spaces, augment and improve existing workflows, and provide a central community structure for health researchers both part of and entering into the Julia ecosystem.
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

@@

% TODO: Cite FunOHDSI
% TODO: Cite OHDSICohortExpressions
% 
% TODO: Cite OHDSI Symposium poster on JuliaHealth
% TODO: Register OHDSI Symposium poster to Zenodo
% TODO: Cite JuliaCon 2023 presentation
% TODO: Register JuliaCon 2023 presentation to Zenodo

@@

% DONE:

@@

% TODO: Mention capr here somehow

@@

% TODO: Add database connector example
% TODO: Reference database connector example

@@

% TODO: Add tutorial to separate repository

@@

\begin{lstlisting}[language=Julia,numbers=none]
import DBConnector: 
  DBConnection
import HealthSampleData: 
  Eunomia
import OMOPCDMCohortCreator as OCC

conn = DBConnection("sqlite", Eunomia())

@@

occ.GenerateTables(conn)
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
using JSON3 
using OHDSICohortExpressions

@@

model = Model(cdm_version = v"5.3.1", 
              cdm_schema = "main",
              vocabulary_schema = "main", 
              results_schema = "main",
              target_schema = "main", 
              target_table = "cohort");

sql = translate(cohort_expression, 
                dialect = :sqlite, 
                model = model, 
                cohort_definition_id = 1);

for query in split(sql, ";")[1:end-1]
  DBI.execute(conn, sub_query)
end
\end{lstlisting}

@@

@@

\begin{lstlisting}[language=Julia,numbers=none]
C = GetCohortSubjects(1, conn)
rename!(C, :subject_id => :person_id)

C_race = occ.GetPatientRace(C.person_id, conn)
C_gender = occ.GetPatientGender(C.person_id, conn)

C_age_group = occ.GetPatientAgeGroup(
  C.person_id, 
  conn; 
)
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
import DataFrames as DF

C_features = DF.outerjoin(C_race,
                          C_gender,
                          C_age_group;
                          on = :person_id, 
                          matchmissing = :equal)

C_features = C_features[:, DF.Not(:person_id)]

C_groups = DF.groupby(C_features, 
                      [
                        :race_concept_id, 
                        :gender_concept_id, 
                        :age_group
                      ])

C_groups = DF.combine(C_groups, DF.nrow => :count)
\end{lstlisting}

@@

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort & Total & Prev. (\%) \\ \hline
        B/AA & F & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        W & F & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        W & F & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        A & M & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        UNK & F & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        B/AA & M & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
\end{table}

@@

\begin{lstlisting}[language=Julia,numbers=none]
import Base:
  Fix2

import OMOPCDMCohortCreator:
  GetPatientRace,
  GetPatientGender,
  GetPatientAgeGroup

CGetPatientRace = Fix2(GetPatientRace, c)
CGetPatientGender = Fix2(GetPatientGender, c)
CGetPatientAgeGroup = Fix2(GetPatientAgeGroup, c)
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
import Chain:
  @chain

CofactorPatients(patients) = @chain patients begin
    CGetPatientRace_fixed
    CGetPatientGender_fixed
    CGetPatientAgeGroup_fixed
end
\end{lstlisting}

\begin{lstlisting}[language=Julia,numbers=none]
CharacterizePatients(patients) = @chain patients begin
    _[:, DF.Not(:person_id)]
    DF.groupby(_, names(_))
    DF.combine(_, DF.nrow => :count)
    occ.ExecuteAudit
end
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
RunStudy(patients) = @chain patients begin
    CofactorPatients
    CharacterizePatients
end
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
import Distributed:
  @distributed

@@

vcat(result...) |> audit_patient_groups
\end{lstlisting}

@@

\begin{lstlisting}[language=Julia,numbers=none]
  GetPatientAgeGroup([1]; age_groupings = [[0, 40], [41, 80]])
\end{lstlisting}

\begin{lstlisting}[language=Julia,numbers=none]
  SELECT
    "PERSON_2"."person_id",
    (CASE 
      WHEN ("PERSON_2"."age" < 41) 
        THEN '0 - 40' 
      WHEN ("PERSON_2"."age" < 81) 
        THEN '41 - 80' END) AS "age_group"
  FROM (
    SELECT
      "PERSON_1"."person_id",
      (2023 - "PERSON_1"."year_of_birth") AS "age"
    FROM "PERSON" AS "PERSON_1"
    WHERE ("PERSON_1"."person_id" IN (1))
  ) AS "PERSON_2"
\end{lstlisting}
